# Visualization {#dataviz}

Module originally written by Shiro Kuriwaki


## Where are we? Where are we headed?

Up till now, you should have covered:

* The R Visualization and Programming primers at  <https://rstudio.cloud/primers/>
* Reading and handling data
* Matrices and Vectors


Today we'll cover:

* Visualization. 
* Data Wrangling

## Check your understanding 

* What does `:` mean in R? What about `==`? `,`?, `!=` , `&`, `|`, `%in% `
* What does `%>%` do
* How do you make a barplot, in base-R and in ggplot?


```{r libraries, message=FALSE, warning=FALSE}
library(dplyr)
library(readr)
library(scales)
library(forcats)
library(ggplot2)
```



## Motivation

In this module, let's jump into using the basic functions of R from an actual Census.

Why care about the Census? The Census is one of the fundamental acts of a government. See the Law Review article by [Persily (2011)](http://cardozolawreview.com/Joomla1.5/content/32-3/Persily.32-3.pdf), "The Law of the Census"^[[Persily, Nathaniel. 2011. "The Law of the Census: How to Count, What to Count, Whom to Count, and Where to Count Them.” _Cardozo Law Review_ 32(3): 755–91.](http://cardozolawreview.com/Joomla1.5/content/32-3/Persily.32-3.pdf)]. The Census is government's primary tool for apportionment (allocating seats to districts), appropriations (allocating federal funding), and tracking demographic change.  See^[[Hochschild, Jennifer L., and Brenna Marea Powell. 2008. "Racial Reorganization and the United States Census 1850–1930: Mulattoes, Half-Breeds, Mixed Parentage, Hindoos, and the Mexican Race." _Studies in American Political Development_ 22(1): 59–96.](https://dash.harvard.edu/bitstream/handle/1/3153295/hoschschild_racialreorganization.pdf?sequence=2)] for example [Hochschild and Powell (2008)](https://dash.harvard.edu/bitstream/handle/1/3153295/hoschschild_racialreorganization.pdf?sequence=2) on how the categorizations of race in the Census during 1850-1930.



## Read data
The first of line code you will write is often to read in data. Here, let's read in a subset of the 2010 Census. This is a 0.01 percent random sample of the entire U.S. Census. 


```{r}
cen10 <- read_csv("input/usc2010_001percent.csv", col_types = cols())
```

The data comes from IPUMS^[[Ruggles, Steven, Katie Genadek, Ronald Goeken, Josiah Grover, and Matthew Sobek. 2015. Integrated Public Use Microdata Series: Version 6.0 dataset](http://doi.org/10.18128/D010.V6.0)], a great source to extract and analyze Census and Census-conducted survey (ACS, CPS) data.


What does this data look like?
```{r}
cen10
```



## Counting
How many people are in your sample. Three ways to count:
```{r}
nrow(cen10)
```


```{r}
dim(cen10)
```


```{r}
length(cen10$year)
```


Multiply this number by the inverse of the sample proportion. Does it equal the total population of the U.S. in 2010^[The 2010 Census reported 308,745,539. [Wikipedia](https://en.wikipedia.org/wiki/2010_United_States_Census)]? 
```{r}
nrow(cen10)*100*100
```

This and all subsequent  tasks involve manipulating and summarizing data, sometimes called "wrangling". There are at least two approaches, "base-R" and "tidyverse functions". The tidyverse functions are add-ons that programmers have built, and appears to be popular.

## Tabulating
Here are two ways to count by group, or to tabulate.

Use the `table` function
```{r}
table(cen10$race)
```


With tidyverse, use `group_by` to group by a variable, and then `summarize` to count. 
```{r}
cen10 %>% 
  group_by(race) %>%
  summarize(count = n())
```

These `tidyverse` commands from the `dplyr` package are newer and not built-in, but they are one of the increasingly more popular ways to wrangle data. 

* 80 percent of your data wrangling needs might be doable with these basic `dplyr` functions: `select`, `mutate`, `group_by`, `summarize`, and `arrange`.
* These verbs roughly correspond to the same commands in SQL, another important language in data science. 
* The `%>%` symbol is a pipe. It takes the thing on the left side and pipes it down to the function on the right side. `cen10 %>% group_by(race)` means take `cen10` and pass it on to the function `group_by`, which will group observations by race. Passing that (`%>%`) to `summarize(count = n())` means to then summarize the grouped output by the variable `count`, which we define by the number of rows `n()`. 


## Visualization
Visualizing data is the key part of communication; good data viz gets the point across^[[Kastellec, Jonathan P., and Eduardo L. Leoni. 2007. "Using Graphs Instead of Tables in Political Science."" _Perspectives on Politics_ 5 (4): 755–71.](http://www.princeton.edu/~jkastell/Tables2Graphs/graphs.pdf)]. 

Two prevalent ways of making graphing are referred to as "base-R" and "ggplot".

### base R

"Base-R"  graphics are graphics that are made with R's default graphics commands. First, let's assign our tabulation to an object, call it `tab_race`

```{r}
tab_race <- table(cen10$race)
``` 

Then put it in the `barplot()` function
```{r, fig.fullwidth = TRUE}
barplot(tab_race)
```


### ggplot
A popular alternative a `ggplot` graphics. `gg` stands for grammar of graphics by Hadley Wickham, and it has a new semantics of explaining graphics in R. Again, first let's set up the data. 

Let's group and count first, like we just did.

```{r, fig.fullwidth = TRUE}
grp_race <- group_by(cen10, race) %>%
  summarize(count = n())
```

We will now plot this grouped set of numbers. The `ggplot()` function takes two main arguments, `data` and `aes`. 

1. First enter a single dataframe from which you will draw a plot.
2. Then enter the `aes`, or aesthetics. This defines which variable in the data the plotting functions should take for pre-set dimensions in graphics. The dimensions `x` and `y` are the most important. We will assign `race` and `count` to them, respectively,
3. After you close `ggplot()` .. add __layers__ by the plus sign. A `geom` is a layer of graphical representation, for example `geom_histogram` renders a histogram, `geom_point` renders a scatter plot. For a barplot, we can use `geom_col()`


```{r, fig.fullwidth = TRUE}
ggplot(data = grp_race, aes(x = race, y = count)) + geom_col()
```


### Improving your graphics
Adjusting your graphics to make the point clear is an important skill. Here is an example of showing the same numbers but with a different design, in a way that aims to maximize the "data-to-ink ratio".


```{r, fig.fullwidth = TRUE}
par(oma = c(1, 10, 1, 1))
barplot(sort(tab_race), # sort numbers
        horiz = TRUE, # flip
        border = NA, # border is extraneous
        xlab = "Number in Race Category", 
        bty = "n", # no box
        las = 1) # alignment of axis labels is horizontal
```

Notice that we applied the `sort()` function to order the bars in terms of their counts. The default ordering of a categorical variable / factor is alphabetical. Alphabetical ordering is uninformative and almost never the way you should order variables.

In ggplot you might do this by:
```{r}
grp_race_ordered <- arrange(grp_race, count) %>% 
  mutate(race = forcats::as_factor(race))

ggplot(data = grp_race_ordered, aes(x = race, y = count)) +
  geom_col() +
  coord_flip() +
  labs(y = "Number in Race Category",
       x = "",
       caption = "Source: 2010 U.S. Census sample")

```




### Cross-tabs
A cross-tab is counting with two types of variables, and is a simple and powerful tool to show the relationship between multiple variables.

```{r}
xtab_race_state <- table(cen10$race, cen10$state)
dim(xtab_race_state)
```

The `barplot` function kindly takes cross-tabs and figures out how to plot them:
```{r}
par(oma = c(2,2,2,2))
barplot(xtab_race_state)
```

What if we care about proportions within states, rather than counts. We want to compare the racial composition of a small state (like Delaware) and a large state (like California). 

One way to transform a table of counts to a table of proportions is the function `prop.table`. Be careful what you want to take proportions of -- this is set by the `margin` argument. In R, the first margin (1) is _rows_ and the second margin (2) is columns.

```{r}
ptab_race_state <- prop.table(xtab_race_state, margin = 2)
dim(ptab_race_state)
```

Now we can plot these numbers. Let's orient them horizontally:
```{r, fig.fullwidth = TRUE, fig.height = 8}
barplot(ptab_race_state, horiz = TRUE, las = 1)
```


To make the same figure with `ggplot()`. 

First, we want a count for each state $\times$ race combination.  So group by those two factors and count how many observations are in each two-way categorization.
```{r}
grp_race_state <- cen10 %>% 
  group_by(race, state) %>% 
  summarize(count = n())
```

Can you tell from the code what `grp_race_state` will look like?

```{r, eval = FALSE}
# run on your own
grp_race_state
```


Now, we want to tell `ggplot2` something like the following: I want bars by state, where heights indicate racial groups. Each bar should be colored by the race. With some googling, you will get something like this 
```{r, fig.height = 8}
ggplot(data = grp_race_state, aes(x = state, y = count,  fill = race)) +
  geom_col(position = "fill") + # the position is dertemined by the fill ae
  scale_fill_brewer(palette = "OrRd", direction = -1) + # choose palette
  coord_flip() + # flip axes
  scale_y_continuous(labels = percent) + # label numbers as percentage
  labs(y = "Proportion of Racial Group within State",
       x = "",
       source = "Source: 2010 Census  sample")
```


### Histograms
When our data is continuous rather than categorical, our first graphics will probably be a histogram.

```{r}
hist(cen10$age, main = "Age Distribution", xlab = "Age")
```


```{r}
ggplot(data = cen10, aes(x = age)) +
  geom_histogram()
```




## Review Exercises
In the time remaining, try the following exercises. Order doesn't matter. 

### Exercise 1: Make your own graphic
Make a graphic from the census data that tells us something interesting about the U.S. 2010 population.
```{r}
# Enter yourself

```


### Exercise 2: Counting CVAP
A issue raised in Persily's article is that the full-count U.S. Census does not record  whether the residents are citizens of the United States^[Here is that argument of his again, more recently in the popular press. ["The Mysterious Number of American Citizens". June 2, 2015.  _POLITICO_](http://www.politico.com/magazine/story/2015/06/the-supreme-courts-big-data-problem-118568)]. Instead, this question is asked in a survey, the American Community Survey. The two are fundamentally different exercises: the Census counts everyone by definition, a survey samples its data. Load the 1 percent sample of the 2015 ACS (`acs2015_1percent.csv`, in the `input` folder) and give an estimate of the proportion of a state's ACS respondents that are reportedly U.S. citizens. 

```{r}
# Enter yourself

```

### Exercise 3: Write your own function
Write your own function that makes some task of data analysis simpler. Ideally, it would be a function that helps you do either of the previous tasks in fewer lines of code. 
```{r}
# Enter yourself

``` 



## Tell us about this session. 
This is day 1 and we covered a lot of material. Some of you might have found this completely new; others not so. Please click through this survey before you leave so we can adjust accordingly on the next few days.
